---
title: "California Congressional Redistricting Analysis"
subtitle: "Comparing 2024 and 2025 District Maps"
format: 
  dashboard:
    orientation: columns
editor: 
  markdown: 
    wrap: 72
---

```{r}
#| include: false
library(tidyverse)
library(plotly)

# Load data
cleaned_data <- read_csv("data/g24_sov_cleaned.csv")

district_2024 <- cleaned_data |>
  group_by(CDDIST) |>
  summarise(
    dem_votes = sum(CNGDEM01, na.rm = TRUE),
    rep_votes = sum(CNGREP01, na.rm = TRUE),
    total_votes = dem_votes + rep_votes,
    .groups = "drop"
  ) |>
  filter(total_votes > 0) |>
  mutate(
    dem_share = dem_votes / total_votes,
    rep_share = rep_votes / total_votes,
    winner = if_else(dem_share > 0.5, "DEM", "REP"),
    map = "2024"
  )

district_2025 <- read_csv("data/district_results_2025.csv") |>
  mutate(
    CDDIST = as.numeric(DISTRICT),
    map = "2025"
  )

# Calculate metrics
compute_eg <- function(dem, rep) {
  total <- dem + rep
  threshold <- floor(total / 2) + 1
  dem_wasted <- if_else(dem > rep, dem - threshold, dem)
  rep_wasted <- if_else(rep > dem, rep - threshold, rep)
  (sum(rep_wasted, na.rm = TRUE) - sum(dem_wasted, na.rm = TRUE)) / sum(total, na.rm = TRUE)
}

mm_2024 <- mean(district_2024$dem_share) - median(district_2024$dem_share)
mm_2025 <- mean(district_2025$dem_share) - median(district_2025$dem_share)
eg_2024 <- compute_eg(district_2024$dem_votes, district_2024$rep_votes)
eg_2025 <- compute_eg(district_2025$dem_votes, district_2025$rep_votes)
```

# Overview {.sidebar}

**Research Question:** Does the 2025 proposed congressional map (AB 604)
increase partisan advantage compared to the 2024 map?

**Data:** 2024 General Election precinct-level voting results applied to
both district maps.

**Key Metrics:** - Mean-Median Score - Efficiency Gap

# Metrics

## Column {width="50%"}

### Mean-Median Score

```{r}
mm_data <- data.frame(
  Map = c("2024", "2025"),
  Score = c(mm_2024, mm_2025)
)

p1 <- ggplot(mm_data, aes(x = Map, y = Score, fill = Map)) +
  geom_col(width = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_fill_manual(values = c("2024" = "#619CFF", "2025" = "#F8766D")) +
  labs(
    title = "Mean-Median Score Comparison",
    y = "Score (positive = pro-DEM bias)",
    x = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "none")

ggplotly(p1)
```

### Interpretation

The mean-median score measures the difference between the mean and
median Democratic vote share across districts.

**2024 Map:** `r round(mm_2024, 4)`\
**2025 Map:** `r round(mm_2025, 4)`\
**Change:** `r round(mm_2025 - mm_2024, 4)`

The change is minimal, suggesting similar central tendencies in vote
distribution.

## Column {width="50%"}

### Efficiency Gap

```{r}
eg_data <- data.frame(
  Map = c("2024", "2025"),
  Gap = c(eg_2024, eg_2025)
)

p2 <- ggplot(eg_data, aes(x = Map, y = Gap, fill = Map)) +
  geom_col(width = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_hline(yintercept = 0.08, linetype = "dotted", color = "red") +
  geom_hline(yintercept = -0.08, linetype = "dotted", color = "red") +
  scale_fill_manual(values = c("2024" = "#619CFF", "2025" = "#F8766D")) +
  labs(
    title = "Efficiency Gap Comparison",
    y = "Gap (negative = pro-DEM advantage)",
    x = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "none")

ggplotly(p2)
```

### Interpretation

The efficiency gap measures the difference in "wasted votes" between
parties. Values beyond ±0.08 indicate significant partisan bias.

**2024 Map:** `r round(eg_2024, 4)`\
**2025 Map:** `r round(eg_2025, 4)`\
**Change:** `r round(eg_2025 - eg_2024, 4)`

The 2025 map shows a **dramatic increase** in pro-Democratic advantage,
with Republicans wasting significantly more votes.

# Vote Distribution

### Democratic Vote Share Distribution

```{r}
combined <- bind_rows(district_2024, district_2025)

p3 <- ggplot(combined, aes(x = dem_share, fill = map)) +
  geom_histogram(alpha = 0.6, position = "identity", bins = 20) +
  geom_vline(data = data.frame(
    map = c("2024", "2025"),
    mean_val = c(mean(district_2024$dem_share), mean(district_2025$dem_share))
  ), aes(xintercept = mean_val, color = map), 
  linetype = "dashed", linewidth = 1) +
  scale_fill_manual(values = c("2024" = "#619CFF", "2025" = "#F8766D")) +
  scale_color_manual(values = c("2024" = "#619CFF", "2025" = "#F8766D")) +
  scale_x_continuous(labels = scales::percent) +
  facet_wrap(~map, ncol = 1) +
  labs(
    title = "Distribution of Democratic Vote Share by District",
    x = "Democratic Vote Share",
    y = "Number of Districts",
    fill = "Map",
    color = "Mean"
  ) +
  theme_minimal()

ggplotly(p3)
```

## Column {width="30%"}

### Summary Statistics

**2024 Map:** - Districts: `r nrow(district_2024)` - Mean Dem Share:
`r scales::percent(mean(district_2024$dem_share), 0.1)` - Median Dem
Share: `r scales::percent(median(district_2024$dem_share), 0.1)` - DEM
Wins: `r sum(district_2024$winner == "DEM")` - REP Wins:
`r sum(district_2024$winner == "REP")`

**2025 Map:** - Districts: `r nrow(district_2025)` - Mean Dem Share:
`r scales::percent(mean(district_2025$dem_share), 0.1)` - Median Dem
Share: `r scales::percent(median(district_2025$dem_share), 0.1)` - DEM
Wins: `r sum(district_2025$winner == "DEM")` - REP Wins:
`r sum(district_2025$winner == "REP")`

### Conclusion

The 2025 proposed map (AB 604) significantly increases Democratic
partisan advantage compared to the 2024 map. While the mean-median score
shows minimal change, the efficiency gap demonstrates a substantial
shift, indicating that the new district boundaries would favor Democrats
in converting votes to seats.
