---
title: Gerrymandering Metrics
format: typst
---

```{r}
library(tidyverse)
cleaned_data <- read_csv("data/g24_sov_cleaned.csv")
```

### 2024 Election Results and the 2024 District Map

```{r}
district_2024 <- cleaned_data |>
group_by(CDDIST) |>
summarise(
dem_votes = sum(CNGDEM01, na.rm = TRUE),
rep_votes = sum(CNGREP01, na.rm = TRUE),
total_votes = dem_votes + rep_votes,
.groups = "drop"
) |>
filter(total_votes > 0) |>
mutate(
dem_share = dem_votes / total_votes,
rep_share = rep_votes / total_votes,
winner = if_else(dem_share > 0.5, "DEM", "REP")
)

district_2024

```

```{r}
mean_dem_share_2024 <- mean(district_2024$dem_share)
median_dem_share_2024 <- median(district_2024$dem_share)
mm_score_2024 <- mean_dem_share_2024 - median_dem_share_2024

mean_dem_share_2024
median_dem_share_2024
mm_score_2024

```

```{r}
compute_efficiency_gap <- function(dem, rep) {
total <- dem + rep
threshold <- floor(total / 2) + 1

dem_wasted <- if_else(dem > rep,
dem - threshold,
dem)
rep_wasted <- if_else(rep > dem,
rep - threshold,
rep)

total_dem_wasted <- sum(dem_wasted, na.rm = TRUE)
total_rep_wasted <- sum(rep_wasted, na.rm = TRUE)
total_votes <- sum(total, na.rm = TRUE)

eg <- (total_rep_wasted - total_dem_wasted) / total_votes
return(eg)
}

eg_2024 <- compute_efficiency_gap(
dem = district_2024$dem_votes,
rep = district_2024$rep_votes
)

eg_2024

```

For the 2024 congressional map, the mean Democratic vote share across districts is 61.6%, and the median is 62.7%, producing a mean–median difference of −0.0116. This slightly negative value suggests a minor pro-Republican structural bias, though the magnitude is small and likely within the range of natural variation.
The efficiency gap, however, is +0.26, indicating that Republican candidates wasted substantially more votes than Democratic candidates. A positive efficiency gap of this size suggests a strong pro-Democratic advantage in translating votes into seats under the 2024 district configuration. In general, values above ±0.08 are considered indicative of significant partisan asymmetry, so this result implies that the 2024 congressional map favors Democrats to a considerable degree.

### 2024 Election Results and the proposed 2025 District Map



