---
title: Gerrymandering Metrics
format: typst
---

```{r}
library(tidyverse)
library(sf)
library(janitor)
cleaned_data <- read_csv("data/g24_sov_cleaned.csv")
```

### 2024 Election Results and the 2024 District Map

```{r}
district_2024 <- cleaned_data |>
group_by(CDDIST) |>
summarise(
dem_votes = sum(CNGDEM01, na.rm = TRUE),
rep_votes = sum(CNGREP01, na.rm = TRUE),
total_votes = dem_votes + rep_votes,
.groups = "drop"
) |>
filter(total_votes > 0) |>
mutate(
dem_share = dem_votes / total_votes,
rep_share = rep_votes / total_votes,
winner = if_else(dem_share > 0.5, "DEM", "REP")
)

district_2024

```

```{r}
mean_dem_share_2024 <- mean(district_2024$dem_share)
median_dem_share_2024 <- median(district_2024$dem_share)
mm_score_2024 <- mean_dem_share_2024 - median_dem_share_2024

mean_dem_share_2024
median_dem_share_2024
mm_score_2024

```

```{r}
compute_efficiency_gap <- function(dem, rep) {
total <- dem + rep
threshold <- floor(total / 2) + 1

dem_wasted <- if_else(dem > rep,
dem - threshold,
dem)
rep_wasted <- if_else(rep > dem,
rep - threshold,
rep)

total_dem_wasted <- sum(dem_wasted, na.rm = TRUE)
total_rep_wasted <- sum(rep_wasted, na.rm = TRUE)
total_votes <- sum(total, na.rm = TRUE)

eg <- (total_rep_wasted - total_dem_wasted) / total_votes
return(eg)
}

eg_2024 <- compute_efficiency_gap(
dem = district_2024$dem_votes,
rep = district_2024$rep_votes
)

eg_2024

```

For the 2024 congressional map, the mean Democratic vote share across districts is 61.6%, and the median is 62.7%, producing a mean–median difference of −0.0116. This slightly negative value suggests a minor pro-Republican structural bias, though the magnitude is small and likely within the range of natural variation. The efficiency gap, however, is +0.26, indicating that Republican candidates wasted substantially more votes than Democratic candidates. A positive efficiency gap of this size suggests a strong pro-Democratic advantage in translating votes into seats under the 2024 district configuration. In general, values above ±0.08 are considered indicative of significant partisan asymmetry, so this result implies that the 2024 congressional map favors Democrats to a considerable degree.

### 2024 Election Results and the proposed 2025 District Map

```{r}
library(sf)
library(tidyverse)

votes_2024 <- read_csv("data/g24_sov_cleaned.csv")

svprec <- st_read("data/mprec/mprec_state_g24_v01_shp.shp")

srprec <- st_read("data/srprec/srprec_state_g24_v01_shp.shp")

```

```{r}
ab604 <- st_read("data/AB604/AB604.shp")
```

```{r}
names(ab604)
names(srprec)
```

```{r}
votes_by_srprec <- votes_2024 |>
  group_by(SVPREC_KEY) |>
  summarise(
    dem_votes = sum(CNGDEM01, na.rm = TRUE),
    rep_votes = sum(CNGREP01, na.rm = TRUE),
    total_votes = sum(TOTVOTE, na.rm = TRUE),
    .groups = "drop"
  )

srprec_with_votes <- srprec |>
  left_join(votes_by_srprec, by = c("SRPREC_KEY" = "SVPREC_KEY"))


if (st_crs(srprec_with_votes) != st_crs(ab604)) {
  cat("Transforming CRS...\n")
  srprec_with_votes <- st_transform(srprec_with_votes, st_crs(ab604))
}

srprec_with_votes <- st_make_valid(srprec_with_votes)
ab604 <- st_make_valid(ab604)

precinct_overlay <- st_intersection(srprec_with_votes, ab604)

precinct_overlay <- precinct_overlay |>
  mutate(piece_area = st_area(geometry))

precinct_overlay <- precinct_overlay |>
  group_by(SRPREC_KEY) |>
  mutate(
    total_precinct_area = sum(piece_area),
    area_proportion = as.numeric(piece_area / total_precinct_area),
    dem_votes_alloc = dem_votes * area_proportion,
    rep_votes_alloc = rep_votes * area_proportion,
    total_votes_alloc = total_votes * area_proportion
  ) |>
  ungroup()


district_2025 <- precinct_overlay |>
  st_drop_geometry() |>
  group_by(DISTRICT) |>
  summarise(
    dem_votes = sum(dem_votes_alloc, na.rm = TRUE),
    rep_votes = sum(rep_votes_alloc, na.rm = TRUE),
    total_votes = sum(total_votes_alloc, na.rm = TRUE),
    .groups = "drop"
  ) |>
  filter(total_votes > 0) |>
  mutate(
    dem_share = dem_votes / total_votes,
    rep_share = rep_votes / total_votes,
    winner = if_else(dem_share > 0.5, "DEM", "REP")
  )
```

```{r}
write_csv(district_2025, "data/district_results_2025.csv")

cat("\nSpatial overlay complete!\n")
cat("Number of 2025 districts:", nrow(district_2025), "\n")
print(district_2025)

precinct_mapping <- precinct_overlay |>
  st_drop_geometry() |>
  select(SRPREC_KEY, DISTRICT, area_proportion, 
         dem_votes_alloc, rep_votes_alloc, total_votes_alloc)

write_csv(precinct_mapping, "data/precinct_to_2025_district.csv")

```

```{r}

district_2025 <- read_csv("data/district_results_2025.csv")
district_2025 <- read_csv("data/district_results_2025.csv") |>
  mutate(DISTRICT = as.numeric(DISTRICT))

mean_dem_share_2025 <- mean(district_2025$dem_share)
median_dem_share_2025 <- median(district_2025$dem_share)
mm_score_2025 <- mean_dem_share_2025 - median_dem_share_2025

mean_dem_share_2025
median_dem_share_2025
mm_score_2025

compute_efficiency_gap <- function(dem, rep) {
  total <- dem + rep
  threshold <- floor(total / 2) + 1
  
  dem_wasted <- if_else(dem > rep,
                        dem - threshold,
                        dem)
  rep_wasted <- if_else(rep > dem,
                        rep - threshold,
                        rep)
  
  total_dem_wasted <- sum(dem_wasted, na.rm = TRUE)
  total_rep_wasted <- sum(rep_wasted, na.rm = TRUE)
  total_votes <- sum(total, na.rm = TRUE)
  
  eg <- (total_rep_wasted - total_dem_wasted) / total_votes
  return(eg)
}

eg_2025 <- compute_efficiency_gap(
  dem = district_2025$dem_votes,
  rep = district_2025$rep_votes
)

eg_2025

cleaned_data <- read_csv("data/g24_sov_cleaned.csv")

district_2024 <- cleaned_data |>
  group_by(CDDIST) |>
  summarise(
    dem_votes = sum(CNGDEM01, na.rm = TRUE),
    rep_votes = sum(CNGREP01, na.rm = TRUE),
    total_votes = dem_votes + rep_votes,
    .groups = "drop"
  ) |>
  filter(total_votes > 0) |>
  mutate(
    dem_share = dem_votes / total_votes
  )

mm_2024 <- mean(district_2024$dem_share) - median(district_2024$dem_share)
eg_2024 <- compute_efficiency_gap(district_2024$dem_votes, district_2024$rep_votes)

mm_2024
mm_score_2025
mm_score_2025 - mm_2024

eg_2024
eg_2025
eg_2025 - eg_2024



```

## Summary

This analysis examined the partisan effects of California's proposed 2025 congressional redistricting map (AB 604) compared to the existing 2024 map, using precinct-level voting data from the 2024 general election.

**Key Findings:**

The 2025 proposed map demonstrates a substantial increase in pro-Democratic partisan advantage compared to the 2024 map:

-   **Mean-Median Score**: The mean-median score changed minimally from -0.012 (2024) to -0.008 (2025), indicating relatively stable central tendency in Democratic vote share across districts.

-   **Efficiency Gap**: The efficiency gap shifted dramatically from 0.26 (2024) to -0.30 (2025), representing a massive increase in Democratic advantage. This 0.56-point swing indicates that under the 2025 map, Republican votes would be significantly more "wasted" (either packed into districts they win by large margins or spread thin in districts they lose).

**Interpretation:**

An efficiency gap magnitude above 0.08 is generally considered indicative of significant partisan gerrymandering. The 2025 map's efficiency gap of -0.30 substantially exceeds this threshold, suggesting the new district boundaries would strongly favor Democratic candidates in translating votes into seats.

This analysis confirms that Proposition 50's redistricting plan would indeed create a more favorable electoral environment for Democrats, consistent with its stated purpose of countering Republican redistricting efforts in states like Texas.
